---
title: "Special data types"
date: "01-14-2020"
output:
  html_document: default
  pdf_document: default
---

Today, we will spend some time talking about some special data types in R.
- factors
- data and time


## Factors


When importing data to R, base R has a burning desire to turn character
information into factor. See for example, `read.table`, and `read.csv`.

```{r}
# to illustrate the issue of `read.csv`, let's write a csv file out of the gapminder dataset
library(gapminder)
library(tidyverse)
write_csv(gapminder, "gapminder.csv")
```

```{r}
# base R function
read.csv("gapminder.csv")
```


```{r}
# readr function
read_csv("gapminder.csv")
```

### Factor inspection
```{r}
levels(gapminder$continent)
nlevels(gapminder$continent)
class(gapminder$continent)
```

```{r}
gapminder %>% count(continent)
fct_count(gapminder$continent)
```

### Dropping unused levels


The number of levels won't change even all the rows corresponding to specific factor level are dropped.
```{r}
h_countries <- c("Egypt", "Haiti", "Romania", "Thailand", "Venezuela")
h_gap <- gapminder %>%
  filter(country %in% h_countries)
nlevels(h_gap$country)
```

```{r}
h_gap$country <- h_gap$country %>%
  fct_drop() %>%
  levels()
```

```{r}
h_gap <- h_gap %>% droplevels()
```



### Change order of the levels

```{r}
## default order is alphabetical
gapminder$continent %>%
  levels()

## order by frequency
gapminder$continent %>%
  fct_infreq() %>%
  levels()

## backwards!
gapminder$continent %>%
  fct_infreq() %>%
  fct_rev() %>%
  levels()
```

Why?

```{r}
ggplot(gapminder) + geom_bar(aes(continent)) + coord_flip()
```

```{r}
ggplot(gapminder) + geom_bar(aes(continent %>% fct_infreq() %>% fct_rev())) +
  xlab("continent") + coord_flip()
```


```{r}
# reorder factor according to values of another variable
gap_asia_2007 <- gapminder %>% filter(year == 2007, continent == "Asia")
ggplot(gap_asia_2007, aes(x = lifeExp, y = fct_reorder(country, lifeExp))) +
  geom_point() + ylab("country")
```


### Change to any order

```{r}
h_gap$country %>% levels()
h_gap$country %>%
  fct_relevel("Romania", "Haiti") %>%
  levels()
```


### Record levels

```{r}
i_gap <- gapminder %>%
  filter(country %in% c("United States", "Sweden", "Australia")) %>%
  droplevels()
i_gap$country %>% levels()
i_gap$country %>%
  fct_recode("USA" = "United States", "Oz" = "Australia") %>%
  levels()
```


## Date and time

```{r}
library(lubridate)
today()
now()  # in UTC (Coordinated Universal Time)
```


```{r}
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")
```
```{r}
ymd_hms("2017-01-31 20:11:59")
mdy_hm("01/31/2017 08:01")

mdy_hm("01/31/2017 08:01", tz = "America/New_York")
```

```{r}
# all the time zone names
OlsonNames
```

```{r}
(t1 <- mdy_hm("01/31/2017 08:01", tz = "America/New_York"))
# convert timezone
with_tz(t1, tzone = "America/Los_Angeles")
# fix a timezone
force_tz(t1, tzone = "America/Los_Angeles")
```

### From individual components
```{r}
library(nycflights13)
flights %>% 
  select(year, month, day, hour, minute)
```

```{r}
(flights_dt <- flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(
    date = make_date(year, month, day),
    time = make_datetime(year, month, day, hour, minute,)))
```

### Get components

```{r}
datetime <- ymd_hms("2016-07-08 12:34:56")

year(datetime)
month(datetime)
month(datetime, label = TRUE)
mday(datetime)
yday(datetime)
wday(datetime)
wday(datetime, label = TRUE, abbr = FALSE)
```


```{r}
flights_dt %>% 
  mutate(wday = wday(time, label = TRUE)) %>% 
  ggplot(aes(x = wday)) +
    geom_bar()
```



# References

https://r4ds.had.co.nz
https://lubridate.tidyverse.org/
https://forcats.tidyverse.org/

