---
title: "API"
output: pdf_document
date: "02-20-2020"
---

```{r}
library(tidyverse)
library(jsonlite)
```

# API

This section lists some examples of public HTTP APIs that publish data in JSON format. These are great to get a sense of the complex structures that are encountered in real world JSON data.

See also https://github.com/public-apis/public-apis for a list of public APIs.

## CitiBike NYC

A single public API that shows location, status and current availability for all stations in the New York City bike sharing imitative. https://www.citibikenyc.com/system-data


```{r}
citibike <- fromJSON("https://gbfs.citibikenyc.com/gbfs/en/station_status.json")
library(lubridate)
as_datetime(citibike$last_updated)
stations <- citibike$data$stations
colnames(stations)
```

```{r}
nrow(stations)
```


## OnWater https://onwater.io/

```{r, cache = TRUE}
# davis
url <- str_glue("https://api.onwater.io/api/v1/results/{lat},{long}", lat = 38.54491, long = -121.74052)
fromJSON(url)
```

```{r, cache = TRUE}
# lake tahoe
url <- str_glue("https://api.onwater.io/api/v1/results/{lat},{long}", lat = 39.0968, long = -120.0324)
fromJSON(url)
```


## Deck of Cards http://deckofcardsapi.com/

It is a very simple API which suffles cards.

```{r, out.width = "20%"}
# get a deck
deck <- fromJSON("https://deckofcardsapi.com/api/deck/new/shuffle/?deck_count=1")
deck_id <- deck$deck_id

# draw two cards
cards <- fromJSON(
    str_glue("https://deckofcardsapi.com/api/deck/{deck_id}/draw/?count={count}", 
              deck_id = deck$deck_id, count = 2), 
    flatten = TRUE)

if (!identical(knitr:::pandoc_to(), "latex")) {
  # don't display the cards in pdf
  knitr::include_graphics(cards$cards$images.svg)
}
```

The paramenters after `?` are called GET parameters. A more formal way to handle GET parameters is to use the
`httr` package. 

```{r}
library(httr)
```


```{r}
endpoint <- str_glue("https://deckofcardsapi.com/api/deck/{deck_id}/draw/", deck_id = deck$deck_id)
r <- GET(endpoint, query = list(count = 3))
json <- content(r, as = "text")
cards <- fromJSON(json)
cards
```



## GeoDataSource https://www.geodatasource.com/

In this secton, we are going to show you how we use an API which requires an API key. API key allows you to use the services the API provides on behalf of yourself.

```{r, eval = FALSE}
r <- GET(
  "https://api.geodatasource.com/cities", 
  query=list(
    key = "THE API KEY COPIED FROM THE WEBSITE",
    lat = 38.5449,
    lng = -121.741))

stop_for_status(r)
json <- content(r, as = "text")
fromJSON(json)
```


There are multiple ways to protect your API key.

- Create a file called `.Renviron` and put your API key into it.

  ```
  GEODATA_KEY="YOUR API KEY"
  ```

```{r}
# you might need to change your working directory and restart R session to make it work
r <- GET(
  "https://api.geodatasource.com/cities", 
  query=list(
    key = Sys.getenv("GEODATA_KEY"),
    lat = 38.5449,
    lng = -121.741))
stop_for_status(r)
json <- content(r, as = "text")
fromJSON(json)
```

- The second appoarch is to make use of the package `keyring`. (PS: this method doesn't work for shiny app)

```{r, eval = FALSE}
# use keyring::key_set to set a password
# only need to do it once, you will be prompted for the API key
keyring::key_set("GEODATA_KEY")
```
```{r, eval = FALSE}
r <- GET(
  "https://api.geodatasource.com/cities", 
  query=list(
    key = keyring::key_get("GEODATA_KEY"),
    lat = 38.5449,
    lng = -121.741))
stop_for_status(r)
json <- content(r, as = "text")
fromJSON(json)
```


# The Guardian News https://open-platform.theguardian.com/

```{r}
search_guardian <- function(text, page = 1) {
  r <- GET(
    "https://content.guardianapis.com/search", 
    query=list(
      `api-key` = Sys.getenv("GUARDIAN_KEY"),
      q = text,
      page = page))
  stop_for_status(r)
  json <- content(r, as = "text", encoding = "UTF-8")
  fromJSON(json)$response
}

response <- search_guardian("coronavirus")
```

```{r}
# number of pages
response$pages

response$results %>% select(webTitle, webPublicationDate)
```

```{r}
search_guardian("coronavirus", 2)$results %>% select(webTitle, webPublicationDate)
```


# Bearer token

Some APIs such as yelp uses Bearer token instead of query string.

First, you will need to register an app on yelp: https://www.yelp.com/developers

```{r}
r <- GET(
  "https://api.yelp.com/v3/businesses/search",
  add_headers(Authorization = paste("Bearer", Sys.getenv("YELP_TOKEN"))),
  query = list(
    location = "Davis"
  )
)
stop_for_status(r)
json <- content(r, as = "text")
fromJSON(json)$businesses %>% select(name)
```

# OAUTH

Some APIs such as twitter, facebook, google, require using OAUTH to authenticate.

> OAuth is an open standard for access delegation, commonly used as a way for Internet users to grant websites or applications access to their information on other websites but without giving them the passwords

In our case, the "application" would be the R code.

I use google as an demo. First, you need to setup an app at https://console.developers.google.com/.
Additionally, you also need to enable the gmail api if you want the manage gmail.


```{r, cache = TRUE}
myapp <- oauth_app(
  "google",
  key = "929233483196-o0ge3pc7q3ec4gbe51ph21rg5tuucbbh.apps.googleusercontent.com",
  secret = Sys.getenv("GOOGLE_SECRET")
)

google_token <- oauth2.0_token(
  oauth_endpoints("google"), 
  myapp,
  scope = c("profile", "email", 
            "https://www.googleapis.com/auth/gmail.readonly"))


google_request <- function(endpoint, query = NULL) {
  r <- GET(endpoint, config(token = google_token), query = query)
  stop_for_status(r)
  json <- content(r, as = "text")
  fromJSON(json)
}

# search mailbox for GeoDataSource
google_request("https://www.googleapis.com/gmail/v1/users/me/messages",
                         query = list(q = "GeoDataSource"))

# Get the title of a specific mail
email <- google_request(
  str_glue("https://www.googleapis.com/gmail/v1/users/me/messages/{thread}", thread = "17060c703052bd61"))
email$payload$headers %>% filter(name == "Subject") %>% select(value)
```

Remark 1: if you just want to manage gmail in R, use https://gmailr.r-lib.org/
Remark 2: if you just want to do google search, use https://serpapi.com/
