---
title: "Web Scrpping"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "02-25-2020"
---

```{r}
library(tidyverse)
library(rvest)
```

# HTML and XML

Here is an example of a simple HTML page:

```html
<!DOCTYPE html>
<html>
<head>
<title>Page Title</title>
</head>
<body>
<h1>This is a Heading</h1>
<p>This is a paragraph.</p>
</body>
</html>
```

Of course, there are a lot of different tags. Go to https://www.w3schools.com/html/ to see some html basics.
Nevertheless, all these tags are predefined so your browser knows what each tag means.

(We only need to only a bit HTML in order to do web scrapping)


# Look at web page source code

It is important to identify the thing that you want to scrape in the webpage. The best way to do it is to use the inspect function in the Chrome browser.

## imdb example

Suppose we want to get the list of most top rated movies from https://www.imdb.com/chart/top/?ref_=nv_mv_250

We see that all movies names are under the `<td>/<a>` nodes. The `<td>` node has a class name `titleColumn`.

```{r}
html <- read_html("https://www.imdb.com/chart/top/?ref_=nv_mv_250")
# it finds all the <td> nodes, but we only need the node with class `titleColumn`
td_nodes <- html %>% html_nodes("td")
# it finds all the <td> nodes with class titleColumn
title_columns <- html %>% html_nodes("td.titleColumn")
# it finds all the <a> nodes within <td> nodes with class titleColumn
a_nodes <- title_columns %>% html_nodes("a")
# html_text to get the values inside the <a> </a> tag
movie_names <- a_nodes %>% html_text()
head(movie_names)
```

Now, we also want to capture the ratings.

```{r}
imdb_ratings <-  html %>% 
  html_nodes("td.ratingColumn.imdbRating") %>% 
  html_nodes("strong") %>% 
  html_text()
```
```{r}
tibble(title = movie_names, rating = imdb_ratings)
```

How if you also want to get the years?

```{r}
years <- html %>% 
  html_nodes("td.titleColumn") %>% 
  html_nodes("span.secondaryInfo") %>% 
  html_text() %>% 
  str_extract("\\d+")
```

```{r}
tibble(title = movie_names, year = years, rating = imdb_ratings)
```

There is also a cute function `html_table`.

```{r}
html %>% html_node("table.chart.full-width") %>% 
  html_table() %>% 
  as_tibble(.name_repair = "unique") %>% 
  transmute(rank_and_title = `Rank & Title`, rating = `IMDb Rating`) %>% 
  separate(rank_and_title, c("rank", "title", "year"), sep = "\n")
```


Now, we want to url link to the movie "The Shawshank Redemption".

```{r}
shawshank_url <- html %>% 
  html_nodes("td.titleColumn") %>% 
  html_nodes("a") %>%
  keep(html_text(.) == "The Shawshank Redemption") %>% 
  html_attr("href")
```

But it is not the complete url, we need to base url.
```{r}
shawshank_full_url <- str_c("https://www.imdb.com/", shawshank_url)
```

Then we could futher scrape things from `shawshank_full_url`.



# Treat HTML as (kind of) XML 

What is XML?

```xml
<?xml version="1.0" encoding="UTF-8"?>
<note>
  <to>Tove</to>
  <from>Jani</from>
  <heading>Reminder</heading>
  <body>Don't forget me this weekend!</body>
</note>
```

See https://www.w3schools.com/xml/ for more XML examples.


In the XML technology, there is a powerful tool called `XPath`.
See https://www.w3schools.com/xml/xpath_syntax.asp for details.


The XPath syntax is quite powerful, we could get the url for Shawshank redeption in almost one line.

```{r}
html %>%
  html_nodes(xpath = "
     //td[@class = 'titleColumn']
     /a[text() = 'The Shawshank Redemption']
     /@href") %>% 
  html_text()
```

We could also select all movies which has rating > 8.5.

```{r}
html %>%
  html_nodes(xpath = "
      //td[@class = 'ratingColumn imdbRating']
      /strong[text() > 8.5]
      /../../td[@class = 'titleColumn']/a/text()") %>% 
  html_text()
```

