---
title: "NoSQL"
output:
  html_document:
    df_print: paged
  pdf_document: 
    latex_engine: xelatex
date: "02-27-2020"
---


SQL vs NoSQL

https://www.guru99.com/sql-vs-nosql.html

| SQL                             | NOSQL                           |
| ----                            | ----                            |
| relational                      | non-relational                  |
| static predefined schema        | dynamic schema                  |
| not good for hierarchical data  | good for hierarchical data      |
| best suited for complex queries | not so good for complex queries |
| verticlly Scalable              | horizontally scalable           |
| table                           | collection              |
| record                          | document                |


Among all the NoSQL databases, MongoDB is the most popular.


# Airbnb sample data

https://docs.atlas.mongodb.com/sample-data/sample-airbnb/


To connect to this MongoDB, you need to either on the campus network or connect via UCDavis VPN.

```{r}
library(tidyverse)
library(mongolite)

m <- mongo("airbnb", db = "data", url = "mongodb://mongouser:secret@alan.ucdavis.edu/data")
```

```{r}
m$count()
```


Query Documents
https://docs.mongodb.com/manual/tutorial/query-documents/


```{r}
# find the first document
m$find("{}", limit = 1)
```


```{r}
# filter fileds
m$find("{}", fields = '{"name": true}')
```

```{r}
# filter keys
m$find('{"price": {"$lt": 20}}', fields = '{"name": true, "price": true}')
```

Check for all the opeartors: https://docs.mongodb.com/manual/reference/operator/query-comparison/#query-selectors-comparison

```{r}
m$find('{"price": {"$lt": 50}, "bedrooms": {"$gte": 3}}',
  fields = '{"name": true, "price": true, "bedrooms": true}'
)
```


```{r}
# sort
m$find('{"bedrooms": {"$gte": 4}}',
  fields = '{"name": true, "price": true}',
  sort = '{"price": -1}',
  limit = 5
)
```

## Query nested fields

```{r}
m$find('{"address.country_code": "US"}',
  fields = '{"name": true, "address.country": 1}'
)
```

## Query an array

```{r}
# find all documents where amenities contain both Oven and Refrigerator
m$find('{"amenities": {"$all": ["Oven", "Refrigerator"]}}', 
       fields = '{"name": true, "amenities": true}')
```

```{r}
# find all documents where amenities contain Oven or Refrigerator or both
m$find('{"amenities": {"$in": ["Oven", "Refrigerator"]}}', 
       fields = '{"name": true, "amenities": true}')
```


## Regular expressions

```{r}
m$find('{"name": {"$regex": "sea", "$options": "i"}}', fields = '{"name": true}')
```

## Iteration

```{r}
it <- m$iterate('{"price": {"$lt": 12}}')
while (TRUE) {
  x <- it$one()
  if (is.null(x)) break
  cat(x$name, "\n")
}
```


## Aggregate

The aggregation pipeline is a framework for data aggregation modeled on the concept of data processing pipelines.

In the following code, there are three stages

- filter the data
- group by and summerize
- arrange data

```{r}
m$aggregate('[
  {"$match": {"$and": [{"bedrooms": {"$ne": null}}, {"bedrooms": {"$ne": 0}}]}},
  {"$group": { "_id": "$bedrooms", "price": { "$avg": "$price" }}},
  {"$sort": { "_id": 1}}
]')
```

User the `$project` stage to rename `_id`.
```{r}
m$aggregate('[
  {"$match": {"$and": [{"bedrooms": {"$ne": null}}, {"bedrooms": {"$ne": 0}}]}},
  {"$group": { "_id": "$bedrooms", "price": { "$avg": "$price" }}},
  {"$project": {
    "_id": false,
    "bedrooms": "$_id",
    "price": true
  }},
  {"$sort": {"bedrooms": 1}}
]')
```

Group by two variables
```{r}
m$aggregate('[
  {"$match": {"$and": [{"bedrooms": {"$ne": null}}, {"bedrooms": {"$ne": 0}}]}},
  {"$group": { 
    "_id": {"property_type": "$property_type", "bedrooms": "$bedrooms"}, 
    "price": { "$avg": "$price" }}},
  {"$project": {
    "_id": false,
    "property_type": "$_id.property_type",
    "bedrooms": "$_id.bedrooms",
    "price": true
  }},
  {"$sort": {"property_type": 1, "bedrooms": 1}}
]')
```

Compute the average price per room grouped by property_type

```{r}
m$aggregate('[
  {"$match": {"$and": [{"bedrooms": {"$ne": null}}, {"bedrooms": {"$ne": 0}}]}},
  {"$project": {
    "property_type": "$property_type",
    "bedrooms": "$bedrooms",
    "price_per_room": {"$divide": ["$price", "$bedrooms"]}
  }},
  {"$group": { 
    "_id": "$property_type", 
    "ave_bedrooms": { "$avg": "$bedrooms" },
    "price_per_room": { "$avg": "$price_per_room" }}
  },
  {"$project": {
    "_id": false,
    "property_type": "$_id",
    "ave_bedrooms": true,
    "price_per_room": true
  }}
]')
```

