---
title: "Special data types"
date: "01-14-2020"
output:
  pdf_document: default
  html_document: default
---

Today, we will spend some time talking about some special data types in R.
- factors (categorial data)
- date and time


## Factors


When importing data to R, base R has a burning desire to turn character
information into factor. See for example, `read.table`, and `read.csv`.

```{r, message = FALSE}
library(tidyverse)
```

```{r, eval = FALSE}
# to illustrate the issue of `read.csv`, let's write a csv file out of the flights dataset
library(nycflights13)
write_csv(flights %>% sample_n(100), "flights.csv")
```

```{r, eval = FALSE}
# base R function, character variables are automatically converted to factors
read.csv("flights.csv")
```


```{r, eval = FALSE}
# tidyverse function, character variables are imported as is
read_csv("flights.csv")
# there are several workarounds,
# 1. we could use `mutate_if` to force the conversion
read_csv("flights.csv") %>%
  mutate_if(is.character, as_factor)
read_csv("flights.csv") %>%
  mutate_if(~ is.character(.) && n_distinct(.) < 50, as_factor)

# 2. we could specify the column types
read_csv("flights.csv", col_types = "iiiddddddfccffddddT")

# 3. use the rstudio import interface
```


### Factor inspection

Let's consider the dataset `gss_cat`: A sample of categorical variables from the General Social survey

```{r}
class(gss_cat$partyid)
levels(gss_cat$partyid)
fct_unique(gss_cat$partyid)
nlevels(gss_cat$partyid)
```

```{r}
gss_cat %>% count(partyid)
gss_cat$partyid %>% fct_count(sort = TRUE)
```

### Combining factors


```{r}
fa <- factor("a")
fb <- factor("b")
fab <- factor(c("a", "b"))

c(fa, fb, fab)  # not what you want!
fct_c(fa, fb, fab)
```


### Dropping unused levels


The number of levels won't change even all the rows corresponding to specific factor level are dropped.
```{r}
gss_cat2 <- gss_cat %>%
  filter(partyid %in% c("Independent", "Strong democrat", "Strong republican"))
nlevels(gss_cat2$partyid)
```

```{r}
# drop unused levels of a specific factor
gss_cat2$partyid <- gss_cat2$partyid %>% fct_drop()
# equivalently
gss_cat2 <- gss_cat2 %>% mutate(partyid = fct_drop(gss_cat2$partyid))

levels(gss_cat2$partyid)
```

```{r}
# drop unused levels for all the factors in a data frame
gss_cat2 <- gss_cat2 %>% droplevels()
```

### Change order of the levels

```{r}
gss_cat$partyid %>%
  levels()

## order by frequency
gss_cat %>% mutate(partyid = partyid %>% fct_infreq())

## backwards!
gss_cat %>% mutate(partyid = partyid %>% fct_infreq() %>% fct_rev())
```

Why?

```{r}
ggplot(gss_cat) + geom_bar(aes(partyid)) + coord_flip()
```

```{r}
ggplot(gss_cat) + geom_bar(aes(partyid %>% fct_infreq() %>% fct_rev())) +
  xlab("partyid") + coord_flip()
```


```{r}
# reorder factor according to values of another variable
gss_cat %>%
  group_by(partyid) %>%
  summarize(mean_age = mean(age, na.rm = TRUE)) %>%
  ggplot(aes(x = mean_age, y = fct_reorder(partyid, mean_age))) +
  geom_point() + ylab("partyid")
```


### Change to any order

```{r}
gss_cat$partyid %>% levels()
gss_cat$partyid %>%
  fct_relevel("Strong republican", "Strong democrat") %>%
  levels()
# use mutate verb to modifly the data frame
gss_cat %>% mutate(partyid = partyid %>% fct_relevel("Strong republican", "Strong democrat"))
```


### Recode levels

```{r}
gss_cat$partyid %>% levels()
gss_cat$partyid %>%
  fct_recode(
    "Independent,near rep" = "Ind,near rep",
    "Independent,near dem" = "Ind,near dem"
  ) %>%
  levels()

# if we need to modifly the data frame, then
gss_cat %>% mutate(partyid = partyid %>%
  fct_recode(
    "Independent,near rep" = "Ind,near rep",
    "Independent,near dem" = "Ind,near dem"
  ))
```

### Collapse levels

```{r}
# collapse small levels automatically
gss_cat %>%
  mutate(partyid = partyid %>% fct_lump(4)) %>%
  ggplot() + geom_bar(aes(partyid))
```

```{r}
# collapse manually
gss_cat %>%
  mutate(partyid = partyid %>% fct_collapse(
    Missing = c("No answer", "Don't know"),
    Rep = c("Strong republican", "Not str republican"),
    Ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    Dem = c("Not str democrat", "Strong democrat"),
    Other = c("Other party")
  )) %>%
  ggplot() + geom_bar(aes(partyid))
```

Remark: there is a bug in forcats v0.4.0 such that the argument `group_other` in `fct_collapse` is malfunction.



## Date and time

```{r}
library(lubridate)
today()
now(tz = "UTC") # in UTC (Coordinated Universal Time)
```

```{r}
# internally, time is stored as the so called "unix time"
# the number of seconds since 1970-01-01 00:00:00 UTC
as.double(now())
as_datetime(1579192970)
```

```{r}
ymd("2020-01-21")
mdy("January 21st, 2020")
dmy("21-Jan-2020")
```
```{r}
ymd_hms("2020-01-21 20:11:59")
mdy_hm("01/21/2020 08:01")

mdy_hm("01/21/2020 08:01", tz = "US/Eastern")
```

```{r}
# OlsonNames() prints all the time zones 
# for the us time zones
OlsonNames() %>% keep(~str_starts(., "US/"))
```

```{r}
(t1 <- mdy_hm("01/21/2020 08:01", tz = "US/Eastern"))
# convert timezone
with_tz(t1, tzone = "US/Pacific")
# fix a timezone
force_tz(t1, tzone = "US/Pacific")
```

### From individual components
```{r}
library(nycflights13)
flights %>%
  select(year, month, day, hour, minute)
```

```{r}
(flights_dt <- flights %>%
  select(year, month, day, hour, minute) %>%
  mutate(
    date = make_date(year, month, day),
    time = make_datetime(year, month, day, hour, minute)
  ))
```

Remark: something was wrong above!

### Get components

```{r}
dt <- ymd_hms("2016-07-08 12:34:56")

year(dt)
month(dt)
month(dt, label = TRUE)
mday(dt)
yday(dt)
wday(dt)
wday(dt, label = TRUE, abbr = FALSE)
hour(dt)
minute(dt)
dst(dt)
```


```{r}
flights_dt %>%
  mutate(wday = wday(time, label = TRUE)) %>%
  ggplot(aes(x = wday)) +
  geom_bar()
```


### Math on data and time

```{r}
# a regular day
nor <- ymd_hms("2018-01-01 01:30:00", tz = "US/Eastern") 
nor + minutes(90)  # literal
nor + dminutes(90)  # duration

# day light saving
gap <- ymd_hms("2018-03-11 01:30:00", tz = "US/Eastern") 
gap + minutes(90)  # literal
gap + dminutes(90)  # duration

# leap year
leap <- ymd("2019-03-01")
leap + years(1)  # literal
leap + dyears(1)  # duration
```


```{r}
t1 <- mdy_hm("01/21/2020 08:31", tz = "US/Eastern")
t2 <- mdy_hm("01/21/2020 12:01", tz = "US/Pacific")
t2 - t1
```

```{r}
# rounding
t1 <- mdy_hm("01/21/2020 08:31", tz = "US/Eastern")
floor_date(t1, unit = "month")
round_date(t1, unit = "month")
ceiling_date(t1, unit = "month")
```


### Interval

```{r}
intvl <- interval(ymd("2019-12-01"), ymd("2020-01-21"))
int_start(intvl)
int_end(intvl)
ymd("2019-12-31") %within% intvl
```


# References

R for Data Science https://r4ds.had.co.nz
Documentation of forcats https://forcats.tidyverse.org/
Documentation of lubridate https://lubridate.tidyverse.org/
