---
title: "Special data types"
date: "01-14-2020"
output:
  html_document: default
  pdf_document: default
---

Today, we will spend some time talking about some special data types in R.
- factors
- data and time


## Factors


When importing data to R, base R has a burning desire to turn character
information into factor. See for example, `read.table`, and `read.csv`.

```{r, message = FALSE}
library(tidyverse)
```

```{r, eval = FALSE}
# to illustrate the issue of `read.csv`, let's write a csv file out of the gapminder dataset
write_csv(gss_cat, "gss_cat.csv")
```

```{r, eval = FALSE}
# base R function, character variables are automatically converted to factors
read.csv("gss_cat.csv")
```


```{r, eval = FALSE}
# readr function, character variables are imported as is
read_csv("gss_cat.csv")
# there are several workarounds,
# 1. we could use `mutate_if` to force the conversion
read_csv("gss_cat.csv") %>% mutate_if(is.character, as_factor)

# 2. we could specify the column types
read_csv("gss_cat.csv", col_types="dfdfffffd")

# 3. use the rstudio import interface
```

### Factor inspection
```{r}
class(gss_cat$partyid)
levels(gss_cat$partyid)
nlevels(gss_cat$partyid)
```

```{r}
gss_cat %>% count(partyid)
fct_count(gss_cat$partyid, sort = TRUE)
```

### Dropping unused levels


The number of levels won't change even all the rows corresponding to specific factor level are dropped.
```{r}
gss_cat2 <- gss_cat %>%
  filter(partyid %in% c("Independent", "Strong democrat", "Strong republican"))
nlevels(gss_cat2$partyid)
```


```{r}
# drop a specific factor
gss_cat2$partyid <- gss_cat2$partyid%>% fct_drop()
levels(gss_cat2$partyid )
```

```{r}
# drop all the factors in a data frame
gss_cat2 <- gss_cat2 %>% droplevels()
```

### Change order of the levels

```{r}
gss_cat$partyid %>%
  levels()

## order by frequency
gss_cat %>% mutate(partyid = partyid %>% fct_infreq())

## backwards!
gss_cat %>% mutate(partyid = partyid %>% fct_infreq() %>% fct_rev())
```

Why?

```{r}
ggplot(gss_cat) + geom_bar(aes(partyid)) + coord_flip()
```

```{r}
ggplot(gss_cat) + geom_bar(aes(partyid %>% fct_infreq() %>% fct_rev())) +
  xlab("partyid") + coord_flip()
```


```{r}
# reorder factor according to values of another variable
gss_cat %>%
  group_by(partyid) %>%
  summarize(mean_age = mean(age, na.rm = TRUE)) %>%
  ggplot(aes(x = mean_age, y = fct_reorder(partyid, mean_age))) +
  geom_point() + ylab("partyid")
```


### Change to any order

```{r}
gss_cat$partyid %>% levels()
gss_cat$partyid %>% fct_relevel("Strong republican", "Strong democrat") %>% levels()
# use mutate verb to modifly the data frame
gss_cat %>% mutate(partyid = partyid %>% fct_relevel("Strong republican", "Strong democrat"))
```


### Recode levelsx

```{r}
gss_cat$partyid %>% levels()
gss_cat$partyid %>%
  fct_recode(
    "Independent,near rep" = "Ind,near rep",
    "Independent,near dem" = "Ind,near dem") %>%
  levels()
# if we need to modifly the data frame, then
gss_cat %>% mutate(partyid = partyid %>%
                     fct_recode(
                       "Independent,near rep" = "Ind,near rep",
                       "Independent,near dem" = "Ind,near dem")
                   )
```

### Collapse levels

```{r}
# collapse automatically
gss_cat %>% 
  mutate(partyid = partyid %>% fct_lump(5)) %>%
  ggplot() + geom_bar(aes(partyid))
```

```{r}
# collapse manually
gss_cat %>% mutate(partyid = partyid %>% fct_collapse(
    Missing = c("No answer", "Don't know"),
    Rep = c("Strong republican", "Not str republican"),
    Ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    Dem = c("Not str democrat", "Strong democrat"),
    Other = c("Other party")
  )) %>%
  ggplot() + geom_bar(aes(partyid))
```
Remark: there is a bug in forcats v0.4.0 such that the argument `group_other` in `fct_collapse` is malfunction.


## Date and time

```{r}
library(lubridate)
today()
now()  # in UTC (Coordinated Universal Time)
```


```{r}
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")
```
```{r}
ymd_hms("2017-01-31 20:11:59")
mdy_hm("01/31/2017 08:01")

mdy_hm("01/31/2017 08:01", tz = "America/New_York")
```

```{r}
# all the time zone names
OlsonNames
```

```{r}
(t1 <- mdy_hm("01/31/2017 08:01", tz = "America/New_York"))
# convert timezone
with_tz(t1, tzone = "America/Los_Angeles")
# fix a timezone
force_tz(t1, tzone = "America/Los_Angeles")
```

### From individual components
```{r}
library(nycflights13)
flights %>% 
  select(year, month, day, hour, minute)
```

```{r}
(flights_dt <- flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(
    date = make_date(year, month, day),
    time = make_datetime(year, month, day, hour, minute)))
```

Remark: something was wrong above!

### Get components

```{r}
datetime <- ymd_hms("2016-07-08 12:34:56")

year(datetime)
month(datetime)
month(datetime, label = TRUE)
mday(datetime)
yday(datetime)
wday(datetime)
wday(datetime, label = TRUE, abbr = FALSE)
```


```{r}
flights_dt %>% 
  mutate(wday = wday(time, label = TRUE)) %>% 
  ggplot(aes(x = wday)) +
    geom_bar()
```



# References

https://r4ds.had.co.nz
https://lubridate.tidyverse.org/
https://forcats.tidyverse.org/

