---
title: "Regular Expression"
output:
  html_document: default
  pdf_document: default
---


# Admin Stuff
- assignment 2 is due today
- introduce `styler`

```{r, message = FALSE}
library(tidyverse)
```

# Regular Expression

## Introduction

A regular expression, `regex` or `regexp` is a sequence of characters that define a search pattern. 

There exists any versions of regular expressions. `stringr` of tidyverse follows the [ICU standard](http://userguide.icu-project.org/strings/regexp) while base R follows the [PCRE](https://www.pcre.org/) standard. See [Comparison of regular-expression engines](https://en.wikipedia.org/wiki/Comparison_of_regular-expression_engines)


We again prefer the tidyverse packages over base functions. 



## Basic concepts

We will use `https://regex101.com/` to illustrate the following. (Choose the Python engine to best mimic ICU standard)

- Boolean "or" - `gray|grey` can match `gray` or `grey`

- Grouping - Parentheses are used to define the scope and precedence of the operators - `gr(a|e)y` can match `gray` and `grey`.

- A quantifier after a token, character or group specifies how often that a preceding element is allowed to occur.

  |quantifier||
  |----|----|
  |?| zero or one occurrences |
  |*|zero or more occurrences|
  |+| one or more occurrences|
  |{n}| extactly n occurrences|
  |{n,}|`n` or more times|
  |{m, n}|`m` or more times but not more than `n`|
  
  Examples:
  - `colou?r` matches both "color" and "colour".
  - `ab*c` matches "ac", "abc", "abbc", "abbbc", and so on
  - `ab+c` matches "abc", "abbc", "abbbc", and so on, but not "ac"

- Wildcard - The wildcard `.` matches any character except a new line.
  
  Examples:
  - `a.c` matches "aac", "abc" and so on.

- Anchors - `^` matches the beginning of a string and `$` matches the end of a string
  
  Examples:
  - `^abc` matches "abc" but not "cabc"
  - `abc$` matches "abc" but not "abcd"

- Bracket expression `[...]` matches a single character that is contained within the brackets
  
  Examples:
    - `[abc]` matches "a", "b", or "c"
    - `[abc123]` matches "a", "b", "c", "1", "2" or "3"
    - `[a-z]` specifies a range which matches any lowercase letter
    from "a" to "z".
    - `[a-zA-Z0-9]` matches all alphanumerics
    - `[\[\]]` matches `[` or `]`
  
- Bracket expression `[^...]` matches a single character that is not contained within the brackets
    
  Examples:
    - `[^abc]` matches any character other than "a", "b", or "c"
    - `[^\]]` matches any `]` character which is not `]`.
  
- Special characters
  
  Honable mentions
  | Pattern | matches|
  |----|---|
  |`\.`| `.`|
  |`\!`| `!`|
  |`\?`| `?`|
  |`\\`| `\`|
  |`\(`| `(`|
  |`\{`| `{`|
  |`\}`| `}`|
  |`\[`| `[`|
  |`\]`| `]`|
  |`\n`| a new line|
  |`\s`| a space or a tab|
  |`\S`| not a space nor a tab|
  |`\d`| a digit|
  |`\D`| a non digit|
  |`\w`| a word character (includes for example CJK chars)|
  |`\W`| a non word character|
  |`\b` | word boundaries|
  ||
  
  
## Escaping characters

In R, these two characters need to be specially treated in double quotes.

  - `"\\"` means a single backslash
  - `"\""` means a double quote


# Package `stringr`

## Manage Strings

```{r}
fruit <- c("apple", "banana", "pear", "pinapple")
str_length(fruit)
# add leading white spaces
str_pad(fruit, 10)
# remove white spaces
str_trim(str_pad(fruit, 10))
# ...
str_trunc(fruit, 5)
```


## Detect Matches

```{r}
fruit <- c("apple", "banana", "pear", "pinapple")
# contains a?
str_detect(fruit, "a")
# starts with a
str_detect(fruit, "^a")
str_starts(fruit, "a")
# ends with a
str_detect(fruit, "a$")
str_ends(fruit, "a")
# contains a, e, i, o or u
str_detect(fruit, "[aeiou]")
```

```{r}
# negate the result
str_detect(fruit, "^p", negate = TRUE)
```

```{r}
# count the number of matches
str_count(fruit, "p")
str_count(fruit, "p{2}")
```

```{r}
# get locations
str_locate(fruit, "a")
str_locate_all(fruit, "a")
```

```{r}
# The pattern variable can also be vectorized
str_locate_all(fruit, c("a", "b", "p", "p"))
```

## Subset Strings

```{r}
fruit <- c("apple", "banana", "pear", "pinapple")
# exact substring from start to end
str_sub(fruit, 1, 3)
str_sub(fruit, -3, -2)
# only select the elements that match
str_subset(fruit, "^a")
# indexs that have matches
str_which(fruit, "^a")
```

```{r}
shopping_list <- c("apples x4", "bag of flour", "bag of sugar", "milk x2")
# numbers
str_extract(shopping_list, "\\d")
# lower case chars
str_extract(shopping_list, "[a-z]+")
# lower case chars of length 1 to 4
str_extract(shopping_list, "[a-z]{1,4}")
# lower case chars of length 1 to 4 with word boundary
str_extract(shopping_list, "\\b[a-z]{1,4}\\b")

str_extract_all(shopping_list, "[a-z]+")
str_extract_all(shopping_list, "[a-z]+", simplify = TRUE)
```


```{r}
strings <- c(" 219 733 8965", 
             "329-293-8753 ", 
             "banana", 
             "239 923 8115 and 842 566 4692", 
             "Work: 579-499-7527", 
             "$1000",
             "Home: 543.355.3679")

phone <- "([2-9][0-9]{2})[- .]([0-9]{3})[- .]([0-9]{4})"

# only the matched pattern
str_extract(strings, phone)
# with subgroups
str_match(strings, phone)
str_match_all(strings, phone)
```


## Mutate Strings

```{r}
str_sub(fruit, 1, 5) <- "APPLE"
fruit
```

```{r}
fruits <- c("one apple", "two pears", "three bananas")
# change the first a, e, i, o and u to -
str_replace(fruits, "[aeiou]", "-")
# change all a, e, i, o and u to -
str_replace(fruits, "[aeiou]", "-")
# apply a function to the matches
str_replace_all(fruits, "[aeiou]", toupper)
# remove all a, e, i, o and u
str_replace_all(fruits, "[aeiou]", "")
str_remove_all(fruits, "[aeiou]")
```


References of the form `\1`, `\2`, etc will be replaced with the contents of the respective matched group

```{r}
str_replace_all(fruits, "([aeiou])", "[\\1]")
```


```{r}
strings <- c("Work: 219 733 8965", 
             "Mobile: 579-499-7527", 
             "Home: 543.355.3679")

phone <- "([2-9][0-9]{2})[- .]([0-9]{3})[- .]([0-9]{4})"

str_replace_all(strings, phone, "(\\1)-\\2-\\3")
```


```{r}
# apply replacement mutliple times
str_replace_all("foobar", c("foo" = "hello", "bar" = "world"))
```

Changes cases

```{r}
str_to_lower(c("one Apple", "tWo BANANAs", "THREE orangeS"))
str_to_upper(c("one Apple", "tWo BANANAs", "THREE orangeS"))
str_to_title(c("one Apple", "tWo BANANAs", "THREE orangeS"))
```


## Join and split

```{r}
str_c("apple", "pie")
str_c(letters, LETTERS)
```

```{r}
str_c(letters, collapse = "")
str_c(letters, LETTERS, collapse = "")

str_flatten(letters)  # faster than str_c(letters, collapse = "") marginally
```

```{r}
fruits <- c(
  "apples and oranges and pears and bananas",
  "pineapples and mangos and guavas"
)

str_split(fruits, " and ")
str_split(fruits, " and ", simplify = TRUE)
str_split(fruits, " and ", n = 2)
str_split(fruits, " and ", n = 3, simplify = TRUE)
# a shorthand for str_split(...n, simplify = TRUE)
str_split_fixed(fruits, " and ", 3)
```
{

## Glue String

```{r}
name <- c("John", "Peter")
age <- c(23, 17)
# get variables from globals
str_glue("{name} is {age}")
```
```{r}
# get variables from arguments
str_glue("{name} is {age}", name = "Anna", age = 43)
```

```{r}
mtcars %>%
  group_by(cyl) %>%
  summarize(m = mean(mpg)) %>% 
  str_glue_data("A {cyl}-cylinder car has an average mpg of {round(m, 2)}.")
```

## Order Strings

```{r}
names <- c("John", "Albert", "Peter", "Charles")
str_order(names)
str_sort(names)
str_sort(names, decreasing = TRUE)
```

```{r}
files <- c("file10", "file2", "file5", "file1")
str_sort(files)
# more natrual order
str_sort(files, numeric = TRUE)
str_order(files, numeric = TRUE)
```



## Pattern interpretation 


Patterns in stringr functions are interpreted as regex in default, you could use `fixed` or `regex` to change the default behavior.

```{r}
strings <- c("abb", "a.b")

str_detect(strings, "a.b")
str_detect(strings, fixed("a.b"))
str_detect(strings, fixed("A.B", ignore_case = TRUE))
```

```{r}
str_match_all("abaa\na", "a+")
str_match_all("abaa\na", regex("^a+", multiline = TRUE))
str_match_all("abaa\na", regex("^A+", ignore_case = TRUE, multiline = TRUE))
```


## An exercise

```{r}
# we need some string
calculus_url <- "https://en.wikipedia.org/wiki/Calculus"
calculus <- read_lines(calculus_url) %>% 
  str_c(collapse = "\n")
```

We want to extract all words from the document.



## More advanced topics of regex


## Reference

- Online regex tester https://regex101.com/
- R for Data Science https://r4ds.had.co.nz/strings.html
