---
title: "JSON"
date: "02-18-2020"
---

# Getting started with JSON

```{r}
library(tidyverse)
library(jsonlite)
```

JavaScript Object Notation (JSON) is an open-standard file format or data interchange format that uses human-readable text to transmit data objects consisting of attribute–value pairs and array data types.

```json
{
  "firstName": "John",
  "lastName": "Smith",
  "isAlive": true,
  "age": 27,
  "address": {
    "streetAddress": "21 2nd Street",
    "city": "New York",
    "state": "NY",
    "postalCode": "10021-3100"
  },
  "phoneNumbers": [
    {
      "type": "home",
      "number": "212 555-1234"
    },
    {
      "type": "office",
      "number": "646 555-4567"
    },
    {
      "type": "mobile",
      "number": "123 456-7890"
    }
  ],
  "children": [],
  "spouse": null
}
```


In JSON, values must be one of the following data types:

- a string
- a number
- an object (JSON object)
- an array
- a boolean
- null


## Parsing JSON from text

```{r}
txt <- "[12, 3, 7]"
x <- fromJSON(txt)
x
```

## Simplification

Simplification is the process where JSON arrays automatically get converted from a list into a more specific R class. The `fromJSON` function has 3 arguments which control the simplification process: `simplifyVector`, `simplifyDataFrame` and `simplifyMatrix`. Each one is enabled by default.

| JSON structure        | Example JSON data                                        | Simplifies to R class | Argument in fromJSON | 
| ----------------------|----------------------------------------------------------|-----------------------|----------------------|
| Array of primitives   | `["Amsterdam", "Rotterdam", "Utrecht", "Den Haag"]`      | Atomic Vector         | simplifyVector       | 
| Array of objects      | `[{"name":"Erik", "age":43}, {"name":"Anna", "age":32}]` | Data Frame            | simplifyDataFrame    | 
| Array of arrays       | `[ [1, 2, 3], [4, 5, 6] ]`                               | Matrix                | simplifyMatrix       |


### Atomic Vectors

When `simplifyVector` is enabled, JSON arrays containing **primitives** (strings, numbers, booleans or null) simplify into an atomic vector:

```{r}
# A JSON array of primitives
json <- '["Mario", "Peach", null, "Bowser"]'
# Simplifies into an atomic vector
fromJSON(json)
```

Without simplification, any JSON array turns into a list: 

```{r}
# No simplification:
fromJSON(json, simplifyVector = FALSE)
```


### Data Frames

When `simplifyDataFrame` is enabled, JSON arrays containing **objects** (key-value pairs) simplify into a data frame:

```{r}
json <-
'[
  {"Name" : "Mario", "Age" : 32, "Occupation" : "Plumber"}, 
  {"Name" : "Peach", "Age" : 21, "Occupation" : "Princess"},
  {},
  {"Name" : "Bowser", "Occupation" : "Koopa"}
]'
mydf <- fromJSON(json)
mydf
```

The data frame gets converted back into the original JSON structure by `toJSON` (whitespace and line breaks are ignorable in JSON).

```{r}
mydf$Ranking <- c(3, 1, 2, 4)
toJSON(mydf, pretty=TRUE)
```

Hence you can go back and forth between dataframes and JSON, without any manual data restructuring.

### Matrices and Arrays

When `simplifyMatrix` is enabled, JSON arrays containing **equal-length sub-arrays** simplify into a matrix (or higher order R array):

```{r}
json <- '[
  [1, 2, 3, 4],
  [5, 6, 7, 8],
  [9, 10, 11, 12]
]'
mymatrix <- fromJSON(json)
mymatrix
```

Again, we can use `toJSON` to convert the matrix or array back into the original JSON structure:

```{r}
toJSON(mymatrix, pretty = TRUE)
```

The simplification works for arrays of arbitrary dimensionality, as long as the dimensions match (R does not support ragged arrays).

```{r}
json <- '[
   [[1, 2], 
    [3, 4]],
   [[5, 6], 
    [7, 8]],
   [[9, 10],
    [11, 12]]
]'
myarray <- fromJSON(json)
myarray[1, , ]
myarray[ , ,1]
```

# API

This section lists some examples of public HTTP APIs that publish data in JSON format. These are great to get a sense of the complex structures that are encountered in real world JSON data.

## CitiBike NYC

A single public API that shows location, status and current availability for all stations in the New York City bike sharing imitative. https://www.citibikenyc.com/system-data


```{r}
citibike <- fromJSON("https://gbfs.citibikenyc.com/gbfs/en/station_status.json")
stations <- citibike$data$stations
colnames(stations)
```

```{r}
nrow(stations)
```


## ProPublica

Below an example from the [ProPublica Nonprofit Explorer API](https://projects.propublica.org/nonprofits/api) where we retrieve the first 10 pages of tax-exempt organizations in the USA, ordered by revenue. The `rbind_pages` function is used to combine the pages into a single data frame.



```{r}
#store all pages in a list first
baseurl <- "https://projects.propublica.org/nonprofits/api/v2/search.json?order=revenue&sort_order=desc"
pages <- list()
for(i in 0:9){
  mydata <- fromJSON(paste0(baseurl, "&page=", i), flatten=TRUE)
  message("Retrieving page ", i)
  pages[[i+1]] <- mydata$organizations
}
#combine all into one
organizations <- rbind_pages(pages)
#check output
nrow(organizations)
```

```{r}
organizations %>% head(10) %>% select(name, city, state)
```



# Reference

jsonlite quick start: https://cran.r-project.org/web/packages/jsonlite/vignettes/json-aaquickstart.html
jsonlite apis: https://cran.r-project.org/web/packages/jsonlite/vignettes/json-apis.html