---
title: "JSON"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "02-18-2020"
---

# Getting started with JSON

```{r}
library(tidyverse)
library(jsonlite)
```

JavaScript Object Notation (JSON) is an open-standard file format or data interchange format that uses human-readable text to transmit data objects consisting of attribute–value pairs and array data types.

```json
{
  "firstName": "John",
  "lastName": "Smith",
  "isAlive": true,
  "age": 27,
  "address": {
    "streetAddress": "21 2nd Street",
    "city": "New York",
    "state": "NY",
    "postalCode": "10021-3100"
  },
  "phoneNumbers": [
    {
      "type": "home",
      "number": "212 555-1234"
    },
    {
      "type": "office",
      "number": "646 555-4567"
    },
    {
      "type": "mobile",
      "number": "123 456-7890"
    }
  ],
  "children": [],
  "spouse": null
}
```


In JSON, values must be one of the following data types:

- a string
- a number
- an object (JSON object)
- an array
- a boolean
- null


## Parsing JSON

```{r}
txt <- "[12, 3, 7]"
x <- fromJSON(txt)
x
```

```{r}
txt <- '{"apple": 1, "banana": 2}'
x <- fromJSON(txt)
x
```


```{r, eval = FALSE}
# from file
fromJSON("somefile.json")
# from the web
fromJSON("https://www.example.com/hello.json")
```


## Simplification

Simplification is the process where JSON arrays automatically get converted from a list into a more specific R class. The `fromJSON` function has 3 arguments which control the simplification process: `simplifyVector`, `simplifyDataFrame` and `simplifyMatrix`. Each one is enabled by default.

| JSON structure        | Example JSON data                                        | Simplifies to R class | Argument in fromJSON | 
| ----------------------|----------------------------------------------------------|-----------------------|----------------------|
| Array of primitives   | `["Amsterdam", "Rotterdam", "Utrecht", "Den Haag"]`      | Atomic Vector         | simplifyVector       | 
| Array of objects      | `[{"name":"Erik", "age":43}, {"name":"Anna", "age":32}]` | Data Frame            | simplifyDataFrame    | 
| Array of arrays       | `[ [1, 2, 3], [4, 5, 6] ]`                               | Matrix                | simplifyMatrix       |


### Atomic Vectors

When `simplifyVector` is enabled, JSON arrays containing **primitives** (strings, numbers, booleans or null) simplify into an atomic vector:

```{r}
# A JSON array of primitives
json <- '["Mario", "Peach", null, "Bowser"]'
# Simplifies into an atomic vector
fromJSON(json)
```

Without simplification, any JSON array turns into a list: 

```{r}
# No simplification:
fromJSON(json, simplifyVector = FALSE)
```


### Data Frames

When `simplifyDataFrame` is enabled, JSON arrays containing **objects** (key-value pairs) simplify into a data frame:

```{r}
json <-
'[
  {"Name" : "Mario", "Age" : 32, "Occupation" : "Plumber"}, 
  {"Name" : "Peach", "Age" : 21, "Occupation" : "Princess"},
  {},
  {"Name" : "Bowser", "Occupation" : "Koopa"}
]'
mydf <- fromJSON(json)
mydf
```

The data frame gets converted back into the original JSON structure by `toJSON` (whitespace and line breaks are ignorable in JSON).

```{r}
mydf$Ranking <- c(3, 1, 2, 4)
toJSON(mydf, pretty = TRUE)
```

Hence you can go back and forth between dataframes and JSON, without any manual data restructuring.

### Matrices and Arrays

When `simplifyMatrix` is enabled, JSON arrays containing **equal-length sub-arrays** simplify into a matrix (or higher order R array):

```{r}
json <- '[
  [1, 2, 3, 4],
  [5, 6, 7, 8],
  [9, 10, 11, 12]
]'
mymatrix <- fromJSON(json)
mymatrix
```

Again, we can use `toJSON` to convert the matrix or array back into the original JSON structure:

```{r}
toJSON(mymatrix, pretty = TRUE)
```

```{r}
fromJSON(json, simplifyMatrix = FALSE)
```

# API

This section lists some examples of public HTTP APIs that publish data in JSON format. These are great to get a sense of the complex structures that are encountered in real world JSON data.

See also https://github.com/public-apis/public-apis for a list of public APIs.

## CitiBike NYC

A single public API that shows location, status and current availability for all stations in the New York City bike sharing imitative. https://www.citibikenyc.com/system-data


```{r}
citibike <- fromJSON("https://gbfs.citibikenyc.com/gbfs/en/station_status.json")
library(lubridate)
as_datetime(citibike$last_updated)
stations <- citibike$data$stations
colnames(stations)
```

```{r}
nrow(stations)
```


## OnWater https://onwater.io/

```{r, cache = TRUE}
# davis
url <- str_glue("https://api.onwater.io/api/v1/results/{lat},{long}", lat = 38.54491, long = -121.74052)
fromJSON(url)
```

```{r, cache = TRUE}
# lake tahoe
url <- str_glue("https://api.onwater.io/api/v1/results/{lat},{long}", lat = 39.0968, long = -120.0324)
fromJSON(url)
```


## Deck of Cards http://deckofcardsapi.com/

It is a very simple API which suffles cards.

```{r, out.width = "20%"}
# get a deck
deck <- fromJSON("https://deckofcardsapi.com/api/deck/new/shuffle/?deck_count=1")
deck_id <- deck$deck_id

# draw two cards
cards <- fromJSON(
    str_glue("https://deckofcardsapi.com/api/deck/{deck_id}/draw/?count={count}", 
              deck_id = deck$deck_id, count = 2), 
    flatten = TRUE)

if (!identical(knitr:::pandoc_to(), "latex")) {
  # don't display the cards in pdf
  knitr::include_graphics(cards$cards$images.svg)
}
```

The paramenters after `?` are called GET parameters. A more formal way to handle GET parameters is to use the
`httr` package. 

```{r}
library(httr)
```


```{r}
endpoint <- str_glue("https://deckofcardsapi.com/api/deck/{deck_id}/draw/", deck_id = deck$deck_id)
r <- GET(endpoint, query = list(count = 3))
json <- content(r, as = "text")
cards <- fromJSON(json)
cards
```



## GeoDataSource https://www.geodatasource.com/

In this secton, we are going to show you how we use an API which requires an API key. API key allows you to use the services the API provides on behalf of yourself.

```{r, eval = FALSE}
r <- GET(
  "https://api.geodatasource.com/cities", 
  query=list(
    key = "THE API KEY COPIED FROM THE WEBSITE",
    lat = 38.5449,
    lng = -121.741))

stop_for_status(r)
json <- content(r, as = "text")
fromJSON(json)
```


There are multiple ways to protect your API key.

- Create a file called `.Renviron` and put your API key into it.

  ```
  GEODATA_KEY="YOUR API KEY"
  ```

```{r}
# you might need to change your working directory and restart R session to make it work
r <- GET(
  "https://api.geodatasource.com/cities", 
  query=list(
    key = Sys.getenv("GEODATA_KEY"),
    lat = 38.5449,
    lng = -121.741))
stop_for_status(r)
json <- content(r, as = "text")
fromJSON(json)
```

- The second appoarch is to make use of the package `keyring`. (PS: this method doesn't work for shiny app)

```{r, eval = FALSE}
# use keyring::key_set to set a password
# only need to do it once, you will be prompted for the API key
keyring::key_set("GEODATA_KEY")
```
```{r, eval = FALSE}
r <- GET(
  "https://api.geodatasource.com/cities", 
  query=list(
    key = keyring::key_get("GEODATA_KEY"),
    lat = 38.5449,
    lng = -121.741))
stop_for_status(r)
json <- content(r, as = "text")
fromJSON(json)
```


# The Guardian News https://open-platform.theguardian.com/

```{r}
search_guardian <- function(text, page = 1) {
  r <- GET(
    "https://content.guardianapis.com/search", 
    query=list(
      `api-key` = Sys.getenv("GUARDIAN_KEY"),
      q = text,
      page = page))
  stop_for_status(r)
  json <- content(r, as = "text", encoding = "UTF-8")
  fromJSON(json)$response
}

response <- search_guardian("coronavirus")
```

```{r}
# number of pages
response$pages

response$results %>% select(webTitle, webPublicationDate)
```

```{r}
search_guardian("coronavirus", 2)$results %>% select(webTitle, webPublicationDate)
```


# Bearer token

Some APIs such as yelp uses Bearer token instead of query string.

First, you will need to register an app on yelp: https://www.yelp.com/developers

```{r}
r <- GET(
  "https://api.yelp.com/v3/businesses/search",
  add_headers(Authorization = paste("Bearer", Sys.getenv("YELP_TOKEN"))),
  query = list(
    location = "Davis"
  )
)
stop_for_status(r)
json <- content(r, as = "text")
fromJSON(json)$businesses %>% select(name)
```

# OAUTH

Some APIs such as twitter, facebook, google, require using OAUTH to authenticate.

> OAuth is an open standard for access delegation, commonly used as a way for Internet users to grant websites or applications access to their information on other websites but without giving them the passwords

In our case, the "application" would be the R code.

I use google as an demo. First, you need to setup an app at https://console.developers.google.com/
For gmail services, you also need to enable the GMail api for your app.

```{r, eval = FALSE}
myapp <- oauth_app(
  "google",
  key = "929233483196-o0ge3pc7q3ec4gbe51ph21rg5tuucbbh.apps.googleusercontent.com",
  secret = Sys.getenv("GOOGLE_SECRET")
)

google_token <- oauth2.0_token(
  oauth_endpoints("google"), 
  myapp,
  scope = c("profile", "email", 
            "https://www.googleapis.com/auth/gmail.readonly"))


google_request <- function(endpoint, query = NULL) {
  r <- GET(endpoint, config(token = google_token), query = query)
  stop_for_status(r)
  json <- content(r, as = "text")
  fromJSON(json)
}

# search mailbox for GeoDataSource
google_request("https://www.googleapis.com/gmail/v1/users/me/messages",
                         query = list(q = "GeoDataSource"))

# Get the title of a specific mail
email <- google_request(
  str_glue("https://www.googleapis.com/gmail/v1/users/me/messages/{thread}", thread = "17060c703052bd61"))
email$payload$headers %>% filter(name == "Subject") %>% select(value)
```

Remark: if you just want to do google search, use this API https://serpapi.com/



# Reference

jsonlite quick start: https://cran.r-project.org/web/packages/jsonlite/vignettes/json-aaquickstart.html
jsonlite apis: https://cran.r-project.org/web/packages/jsonlite/vignettes/json-apis.html

