---
title: "Interactive visualization"
date: "01-21-2020"
---

```{r, message = FALSE}
library(tidyverse)
library(plotly)
```

# plotly vs ggplot2

```{r}
mpg %>%
  mutate(cyl = as_factor(cyl)) %>% 
  ggplot(aes(displ, hwy, color = cyl)) + geom_point()
```

```{r, message = FALSE}
mpg %>% 
  plot_ly(x = ~displ, y = ~hwy, color =  ~factor(cyl))
```

Pros:
- Interactive 
- Plotly handles multiple wide data columns (ggplot2 requies long format)
- Plotly works for Python, Matlab, and Excel, among other languages
- Easy layout customization
- 3D charts

Cons:
- Doesn't work very well with pdf
- Facet wrapping is a bit complicated compared with ggplot2
- adding legend title is difficult


```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(data = mutate(mpg, cyl = NULL), color = "grey75") +
  geom_point() +
  facet_wrap(vars(cyl))
```



## `ggplotly` function for ggplot2 users

```{r message=FALSE}
p <- mpg %>%
  mutate(cyl = as_factor(cyl)) %>%
  ggplot(aes(displ, hwy, color = cyl)) + geom_point()
ggplotly(p)
```


## We won't learn (much of)

- ggplot2 (or should we?)

- HTML, SVG, CSS, JavaScript

- d3.js (R package r2d3)


## Scatter plots + lines

```{r message=FALSE}
p <- economics %>%
  sample_n(n()) %>%
  plot_ly(x = ~date, y = ~psavert)
p %>% add_paths()  # using the order of the data frame 
p %>% add_lines()
```


```{r}
library(lubridate)
econ <- economics %>%
  mutate(yr = year(date), mnth = month(date))

# One trace (more performant, but less interactive)
econ %>%
  group_by(yr) %>%
  plot_ly(x = ~mnth, y = ~uempmed) %>%
  add_lines(text = ~yr)

# Multiple traces (less performant, but more interactive)
econ %>% 
  mutate(yr = ordered(yr)) %>% 
  plot_ly(x = ~mnth, y = ~uempmed) %>%
  add_lines(color = ~yr)
```

### Use Canvas rather then SVG for large dataset

```{r}
plot_ly(econ, x = ~mnth, y = ~uempmed) %>%
  add_lines(color = ~ordered(yr)) %>%
  toWebGL()
```


### Markers

- alpha blending to see replicates

```{r, message = FALSE}
mpg %>% 
  plot_ly(x = ~cty, y = ~hwy) %>% 
  add_markers(alpha = 0.2)
```


- colors for grouping

```{r, message = FALSE}
mtcars %>% 
  plot_ly(x = ~disp, y = ~mpg) %>% 
  add_markers(color = ~factor(cyl))
```

- symbols

```{r, message = FALSE}
mtcars %>% 
  plot_ly(x = ~disp, y = ~mpg) %>% 
  add_markers(symbol = ~factor(cyl))
```

## Error bar plot


```{r, message = FALSE}
mpg %>% 
  group_by(cyl) %>%
  summarize(mhwy = mean(hwy), se = sd(hwy)/sqrt(n())) %>%
  plot_ly(x = ~mhwy, y = ~factor(cyl)) %>%
  add_markers(error_x = ~list(value = se)) %>%
  layout(xaxis = list(title = "mean hwy"), yaxis = list(title = "cyl"))
```

## Dumbbell Plot

```{r, message = FALSE}
mpg %>%
  group_by(model) %>%
  summarize(c = mean(cty), h = mean(hwy)) %>%
  mutate(model = forcats::fct_reorder(model, c)) %>%
  plot_ly() %>%
  add_segments(
    x = ~c, y = ~model,
    xend = ~h, yend = ~model, 
    color = I("gray"), showlegend = FALSE
  ) %>%
  add_markers(
    x = ~c, y = ~model, 
    color = I("blue"), 
    name = "mpg city"
  ) %>%
  add_markers(
    x = ~h, y = ~model, 
    color = I("red"),
    name  = "mpg highway"
  ) %>%
  layout(xaxis = list(title = "Miles per gallon"))
```



## Histograms


```{r, message = FALSE}
mpg %>%
  plot_ly(x= ~hwy, color = ~factor(cyl)) %>% 
  add_histogram(histnorm = "", alpha = 0.7) %>%  # histnorm could be "", "probability", "density" and "probability density"
  layout(barmode = "overlay")  # barmode could be "overlay", "stack" and "group"
```

```{r}
names(relig_income) %>% paste0(collapse = "\n") %>% cat
``` 



## Bar plots

```{r, message = FALSE}
mpg %>%
  mutate(cyl = as_factor(cyl)) %>% 
  count(drv, cyl) %>%
  plot_ly(x = ~drv, y = ~n, color = ~cyl) %>% 
  add_bars() %>% 
  layout(barmode = "stack")
```


```{r, message = FALSE}
# work with wide format directly
relig_income %>% 
  mutate(religion = as_factor(religion)) %>%
  plot_ly(y = ~religion) %>% 
  add_bars(~`<$10k`, name = "<$10k") %>%
  add_bars(~`$10-20k`, name = "$10-20k") %>%
  add_bars(~`$20-30k`, name = "$20-30k") %>%
  add_bars(~`$30-40k`, name = "$30-40k") %>%
  add_bars(~`$40-50k`, name = "$40-50k") %>%
  add_bars(~`$50-75k`, name = "$50-75k") %>%
  add_bars(~`$75-100k`, name = "$75-100k") %>%
  add_bars(~`$100-150k`, name = "$100-150k") %>%
  add_bars(~`>150k`, name = ">150k") %>%
  add_bars(~`Don't know/refused`, name = "Don't know/refused") %>%
  layout(xaxis = list(title = "count"), barmode = "stack")
```

```{r, message = FALSE}
# may be easier with `pivot_longer`?
relig_income %>% 
  mutate(religion = as_factor(religion)) %>%
  pivot_longer(-religion, names_to = "income", values_to = "count") %>%
  mutate(income = fct_inorder(income)) %>% 
  plot_ly(x = ~count, y = ~religion, color = ~income) %>% 
  add_bars() %>%
  layout(barmode = "stack")
```


## Box plots

```{r, message = FALSE}
mpg %>%
  mutate(cyl = as_factor(cyl)) %>% 
  plot_ly(x = ~drv, y = ~hwy, color = ~cyl) %>% 
  add_boxplot() %>% 
  layout(boxmode = "group")
```

## Sunburst

```{r, message = FALSE}
library(gapminder)
gapminder2007 <- gapminder %>%
  filter(year == 2007)
gapminder2007 %>% 
  select(children = country, parents = continent, pop = pop) %>%
  bind_rows(
    gapminder2007 %>%
      group_by(continent) %>%
      summarize() %>% 
      transmute(children = continent, parents = "World", pop = 0)
  ) %>%
  add_row(children = "World", parents = "", pop = 0) %>% 
  plot_ly(labels = ~children, parents = ~parents, values = ~pop, type = "sunburst")
```

## Map

While plotly has map plotting feature, we will use a more popular map library `leaflet`.

```{r}
library(leaflet)
leaflet() %>% setView(lng = -121.7405, lat = 38.5449, zoom = 13) %>% 
  addTiles()
```


```{r}
quakes %>% 
  filter(mag > quantile(mag, 0.95)) %>% 
  leaflet() %>%
  addProviderTiles(providers$Wikimedia) %>% 
  addMarkers(~long, ~lat, label = ~mag)
```

### Choropleths

To draw a choropleth, we first need the map data. There is a package `tigris` to download us map data from census.

```{r, message = FALSE}
library(leaflet)
library(tigris)
states <- states(cb = TRUE)  # lowest resolution us map
class(states)
```

```{r, message = FALSE}
starbucks <- read_csv("starbucks.csv")
(starbucks <- starbucks %>%
  count(Province) %>%
  rename(state = Province, total = n))
```


```{r}
states_starbucks <- states %>% 
  geo_join(starbucks, "STUSPS", "state") %>%
  subset(!is.na(total))
pal <- colorNumeric("Greens", domain=states_starbucks$total)
```

```{r, message = FALSE}
states_starbucks %>%
  leaflet() %>%
  setView(lng = -100, lat = 40, zoom = 4) %>% 
  addProviderTiles(providers$Wikimedia) %>% 
  addPolygons(fillColor = ~pal(total), fillOpacity = 0.7, weight = 1, smoothFactor = 0.2) %>% 
  addLegend(pal = pal, values = states_starbucks$total, position = "bottomright", title = "Starbucks")
```



## 3D charts and its family

```{r, message = FALSE}
mpg %>%
  mutate(cyl = as_factor(cyl)) %>%
  plot_ly(x = ~cty, y = ~hwy, z = ~cyl) %>%
  add_markers(color = ~cyl)
```

```{r, message = FALSE}
x <- seq_len(nrow(volcano)) + 100
y <- seq_len(ncol(volcano)) + 500
plot_ly(x = ~x, y = ~y, z = ~volcano) %>% 
  add_surface(colorscale = "Earth")
```

```{r}
# heatmap
plot_ly(x = ~x, y = ~y, z = ~volcano) %>% 
  add_heatmap(colorscale = "Earth")
```

```{r}
plot_ly(x = ~x, y = ~y, z = ~volcano) %>% 
  add_contour(colorscale = "Earth")
```



## Animating views

```{r}
df <- tibble(
  x = c(1,2,1), 
  y = c(1,2,1), 
  t = c(1,2,3)
)

df %>%
  plot_ly(x = ~x, y = ~y, frame = ~t, showlegend = F)
```


```{r, message = FALSE}
library(gapminder)
(p <- gapminder %>%
  plot_ly(x = ~lifeExp, y = ~gdpPercap, size = ~pop, color = ~continent, frame = ~year) %>%
  layout(yaxis = list(type = "log" )))
```



## A more powerful (therefore difficult to master) graphic library `r2d3`

`r2d3` is an R binding to the famous javascript library `d3.js`. Visit https://github.com/d3/d3/wiki/Gallery to see
some of the things that `d3.js` is able to create.


# References

- Plotly official website: https://plot.ly/r/
- Carson Sievert, Interactive web-based data visualization with R, plotly, and shiny https://plotly-r.com/