---
title: "Database and SQL"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "02-11-2020"
---

```{r, message = FALSE}
library(tidyverse)
library(DBI)
```

```{r, echo = FALSE}
# to suppress the object not found error when kniting
sakila_psql <- NULL
```


What is a database? It is what google says

> a structured set of data held in a computer, especially one that is accessible in various ways.

A relational database is a type of database that stores and provides access to data points that are related to one another. Relation databases are administrated by a Relational Database Management System (RDBMS). The data in RDBMS is stored in database objects called tables. A table is a collection of related data entries and it consists of columns and rows.

There are many RDBMS
- MySQL (owned by Oracle)
- PostgreSQL (open source)
- SQL Server (microsoft)
- SQLite (open source, single file)

What is SQL? Structured Query Language (or SQL) is a standard language for accessing and manipulating relational databaes.
However, each RDMBS may have their own extension of the SQL language and their implmentation may vary too.


## Connect to a databse

We are going to use a popular database called Sakila https://www.jooq.org/sakila.

The Sakila database is a nicely normalised schema modelling a DVD rental store, featuring things like films, actors, film-actor relationships, and a central inventory table that connects films, stores, and rentals.

In the following, we are going to use both `sqlite` and `postgresql`.


### SQLite

The database is called `sakila.sqlite`. You could either git clone from lectures repo or download using the code

```{r}
if (!file.exists("sakila.sqlite") || file.size("sakila.sqlite") == 0) {
  download.file(
    "https://github.com/UCDavis-STA-141B-Winter-2020/sta141b-lectures/raw/master/02-11/sakila.sqlite", 
    destfile = "sakila.sqlite")
}
```

The file format is `.sqlite` which is one of the very common relational database formats, espeically for simple problems.

```{r}
sakila_lite <- dbConnect(RSQLite::SQLite(), dbname = "sakila.sqlite")
sakila_lite %>% dbListTables()
```

### Postgresql

I have also uploaded the Sakila database to a postgres server owned by the department. (You'll need to either on the campus or over UCD vpn to connect to it)

```{r, eval = FALSE}
sakila_psql <- dbConnect(RPostgres::Postgres(),
  dbname = "sakila",
  user = "psqluser", password = "secret", host = "alan.ucdavis.edu"
)
sakila_psql %>% dbListTables()
```


## How not to use SQL?

`dplyr` provides an excellent interface for users without any SQL background to query databases.

```{r}
# number of rental transactions
sakila_lite %>%
  tbl("rental") %>%
  count() %>%
  collect()
```

`sakila_lite %>% tbl("rental")` creates a virtual table rather loading the whole table into memory.

```{r}
sakila_lite %>%
  tbl("rental") %>%
  class()
```

```{r}
sakila_lite %>%
  tbl("rental") %>%
  colnames()
```


## Sakila queries

https://datamastery.gitlab.io/exercises/sakila-queries.html

- Which actors have the first name Scarlett?

```{r}
sakila_lite %>%
  tbl("actor") %>%
  filter(str_to_lower(first_name) == str_to_lower("Scarlett")) %>%
  collect()
```

Suppose we want to make the result a bit more beautiful.

```{r}
sakila_lite %>%
  tbl("actor") %>%
  filter(str_to_lower(first_name) == str_to_lower("Scarlett")) %>%
  collect() %>%
  mutate(first_name = str_to_title(first_name), last_name = str_to_title(last_name))
```

Note: SQLite doesn't support transforming title case but Postgresql does.

```{r, eval = FALSE}
sakila_psql %>%
  tbl("actor") %>%
  filter(str_to_lower(first_name) == str_to_lower("Scarlett")) %>%
  mutate(first_name = str_to_title(first_name), last_name = str_to_title(last_name)) %>% 
  collect()
```

- Which actors have the last name Johansson?

```{r}
sakila_lite %>%
  tbl("actor") %>%
  filter(str_to_lower(last_name) == "johansson") %>%
  collect()
```

- How many distinct actors last names are there?

```{r}
sakila_lite %>%
  tbl("actor") %>%
  summarize(n = n_distinct(last_name)) %>%
  collect()
```

- Which last names are not repeated?

```{r}
sakila_lite %>%
  tbl("actor") %>%
  count(last_name) %>%
  filter(n == 1) %>%
  collect()
```

- Which last names appear more than once?

```{r}
sakila_lite %>%
  tbl("actor") %>%
  count(last_name) %>%
  filter(n > 1) %>%
  collect()
```


- Which actor has appeared in the most films?

```{r}
sakila_lite %>%
  tbl("film_actor") %>%
  count(actor_id) %>%
  arrange(desc(n)) %>%
  head(1) %>%
  inner_join(tbl(sakila_lite, "actor"), by = "actor_id") %>%
  collect()
```

- What is that average running time of all the films in the sakila DB?

```{r}
sakila_lite %>%
  tbl("film") %>%
  summarize(m = mean(length)) %>%
  collect()
```

- What is the average running time of films by category?

```{r}
sakila_lite %>%
  tbl("film") %>%
  left_join(tbl(sakila_lite, "film_category"), by = "film_id") %>%
  group_by(category_id) %>%
  summarize(mean_length = mean(length)) %>%
  left_join(tbl(sakila_lite, "category"), by = "category_id") %>%
  select(name, mean_length) %>%
  collect()
```

- Is 'Unforgiven Zoolander' available for rent from Store 1?

```{r}
uz <- sakila_lite %>%
  tbl("film") %>%
  filter(str_to_lower(title) == str_to_lower("Unforgiven Zoolander")) %>%
  select(film_id)
all_inventories_of_store1 <- sakila_lite %>%
  tbl("inventory") %>%
  filter(store_id == 1) %>%
  select(film_id, inventory_id, store_id)
not_yet_returned <- sakila_lite %>%
  tbl("rental") %>%
  filter(is.na(return_date)) %>%
  select(inventory_id)
uz %>%
  inner_join(all_inventories_of_store1, by = "film_id") %>%
  anti_join(not_yet_returned) %>%
  count() %>%
  collect()
```

## SQL

We just see some example queries of a relational database. Beind the scene, we are using a language called SQL. For example, in the last query, the SQL used is

```{r}
uz %>%
  left_join(all_inventories_of_store1) %>%
  anti_join(not_yet_returned) %>%
  count() %>%
  show_query()
```

Why learning SQL when there is dplyr?

- SQL is everywhere (used in python, php, etc..)
- dplyr magics only read, desnot write
- Job interviews


In R, a sql query can be made by using `dbGetQuery`

```{r}
sakila_lite %>%
  dbGetQuery("SELECT COUNT() AS `n` FROM `rental`")
```

We could also make SQL query by sql block. In here, we are using the connection `sakila_lite`. The result will be printed directly.

```{sql connection = sakila_lite}
SELECT COUNT() AS `n` FROM `rental`;
```

In we need the output, set `output.var` to `rental_count`

```{sql connection=sakila_lite, output.var = "rental_count"}
SELECT COUNT() AS `n` FROM `rental`;
```

The output could be later used in R blocks
```{r}
rental_count
```

For comparsion, in Python, we use
```{python, eval = FALSE}
import sqlite3
sakila_lite = sqlite3.connect('sakila.sqlite')
c = sakila.cursor()
c.execute("SELECT COUNT() AS `n` FROM `rental`")
c.fetchall()
```

SQLite supports both double quotes (which is the standard) and backticks to quote identifiers. Backticks are used in another popular database MySQL. Double quotes are used in Postgresql. It is always a good practice to quote the identifers.


```{sql connection = sakila_lite, eval = FALSE}
-- MySQL style
SELECT COUNT() AS `n` from `actor`;
```

```{sql connection = sakila_psql, eval = FALSE}
-- The standard way
SELECT COUNT() AS "n" from "actor";
```

### SELECT

The SELECT statement is pretty much the `select()` function in `dplyr`.

```{sql connection = sakila_lite}
SELECT "last_name" FROM "actor";
```


```{sql connection = sakila_lite}
SELECT LOWER("last_name") AS "family_name" FROM "actor";
```

For comparision,
```{r}
sakila_lite %>%
  tbl("actor") %>%
  transmute(family_name = str_to_lower(last_name))
```


```{sql connection = sakila_lite}
SELECT * FROM "actor";
```

```{sql connection = sakila_lite}
SELECT "rental_id", "last_update" FROM "rental";
```


### ORDER BY Clause

It is equivalent to `arrange()` in `dplyr`

```{sql connection = sakila_lite}
SELECT * FROM "actor" ORDER BY "last_name";
```

```{sql connection = sakila_lite}
SELECT * FROM "actor" ORDER BY "last_name" DESC;
```


### DISTINCT

DISTINCT operator to remove duplicates from a result set. It is equivalent to `distinct()` function in `dplyr`.

```{sql connection = sakila_lite}
SELECT DISTINCT "last_name" FROM "actor";
```

```{r}
sakila_lite %>%
  tbl("actor") %>%
  distinct(last_name)
```

### LIMIT

```{sql connection = sakila_lite}
SELECT * FROM "actor" LIMIT 2;
```

```{r}
sakila_lite %>%
  tbl("actor") %>%
  head(2)
```

### WHERE

It is equivalent to `filter()` in `dplyr`.

- SQLite (and MySQL) allows double qoutes to quote string values but it is actually not the SQL standard.
In SQL standard, strings are quoted in single quotes.

- In SQL standard, we should use `=` for comparison, but not `==`.

```{sql connection = sakila_lite}
SELECT * FROM "film" WHERE "rating" = 'PG' AND "length" > 90;
```

```{r}
sakila_lite %>%
  tbl("film") %>%
  filter(rating == "PG" && length > 90)
```

- The `IN` operator

```{sql connection = sakila_lite}
SELECT * FROM "film" WHERE "rating" IN ('PG', 'PG-13');
```

```{r}
sakila_lite %>%
  tbl("film") %>%
  filter(rating %in% c("PG", "PG-13"))
```

- The `LIKE` operator

```{sql connection = sakila_lite}
SELECT "title" FROM "film" WHERE "title" LIKE '%victory%';
```

Remark: in SQLite, the `LIKE` operator is case insensitive. However, it is not the case for other DBs.


```{sql connection = sakila_psql, eval = FALSE}
SELECT "title" FROM "film" WHERE "title" LIKE '%victory%';
```

In Postgres, there is a `ILIKE` (case insensitive `LIKE`) operator
```{sql connection = sakila_psql, eval = FALSE}
SELECT "title" FROM "film" WHERE "title" ILIKE '%victory%';
```


### SQL interpolation

```{r}
rating <- "PG"
x <- 90
```

```{sql connection = sakila_lite}
SELECT * FROM "film" WHERE "rating" == ?rating AND "length" > ?x;
```


### CASE

Similar to `case_when()` in `dplyr`.

```{sql connection = sakila_lite}
SELECT 
  "film_id", 
  "title", 
  CASE 
    WHEN "length" < 60 THEN 'short'
    WHEN "length" < 90 THEN 'mid'
    ELSE 'long'
  END "length"
FROM "film";
```

### JOIN operations


- Inner Join - selects records that have matching values in both tables.

```{sql connection = sakila_lite}
SELECT a."inventory_id", b."customer_id"
  FROM "inventory" a
  JOIN "rental" b ON a."inventory_id" = b."inventory_id"
  ORDER BY a."inventory_id"
```

We could also join a more sophisticated subquery.

```{sql connection = sakila_lite}
SELECT a."inventory_id", b."customer_id"
  FROM "inventory" a 
  JOIN (
    SELECT * FROM "rental" WHERE "return_date" IS NULL
  ) b ON a."inventory_id" = b."inventory_id"
  ORDER BY a."inventory_id"
```


- Left Join - returns all records from the left table, and the matched records from the right table

```{sql connection = sakila_lite}
SELECT a."inventory_id", b."customer_id"
  FROM "inventory" a 
  LEFT JOIN (
    SELECT * FROM "rental" WHERE "return_date" IS NULL
  ) b ON a."inventory_id" = b."inventory_id"
  ORDER BY a."inventory_id"
```

- Full Join - returns all records when there is a match in left or right table records.

SQLite doesn't support full join.

```{sql connection = sakila_lite, eval = FALSE}
SELECT a."inventory_id", b."customer_id"
  FROM "inventory" a 
  FULL JOIN (
    SELECT * FROM "rental" WHERE "return_date" IS NULL
  ) b ON a."inventory_id" = b."inventory_id"
  ORDER BY a."inventory_id"
```

- Semi Join - return all records in the left table which has a match in the right table.

```{sql connection = sakila_lite}
SELECT "inventory_id"
  FROM "inventory" i
  WHERE EXISTS (
    SELECT * FROM "rental" r
      WHERE r."inventory_id" = i."inventory_id" AND "return_date" IS NULL
  )
```


- Anti Join - remove all records in the left table which has a match in the right table.

```{sql connection = sakila_lite}
SELECT "inventory_id"
  FROM "inventory" i
  WHERE NOT EXISTS (
    SELECT * FROM "rental" r
      WHERE r."inventory_id" = i."inventory_id" AND "return_date" IS NULL
  )
```


### Aggregate Functions

- AVG – calculate the average value of a set.
- COUNT – return the number of items in a set.
- SUM – return the sum all or distinct items of a set.
- MAX – find the maximum value in a set.
- MIN – find the minimum value in a set.

```{sql connection = sakila_lite}
SELECT AVG("length") as "avg_length" FROM "film"
```


### Group By

```{sql connection = sakila_lite}
SELECT "rating", "rental_duration", AVG("length") AS "avg_length" 
  FROM "film" GROUP BY "rating";
```

```{sql connection = sakila_lite}
SELECT "rating", "rental_duration", AVG("length") AS "avg_length" 
  FROM "film" GROUP BY "rating", "rental_duration";
```

### SET Operators

- UNION and UNION ALL – combine result set of two or more queries into a single result set using the UNION and UNION ALL operators.
- INTERSECT  – return the intersection of two or more queries using the INTERSECT operator.
- EXCEPT – subtract a result set from another result set using the EXCEPT operator

```{sql connection = sakila_lite}
SELECT "film_id" FROM "film" where "film_id" <= 3
UNION
SELECT "film_id" FROM "film" where "film_id" <= 4
```
`UNION ALL` doesn't remove duplications

```{sql connection = sakila_lite}
SELECT "film_id" FROM "film" where "film_id" <= 3
UNION ALL
SELECT "film_id" FROM "film" where "film_id" <= 4
```

```{sql connection = sakila_lite}
SELECT "film_id" FROM "film" where "film_id" <= 5
INTERSECT
SELECT "film_id" FROM "film" where "film_id" >= 2;
```

```{sql connection = sakila_lite}
SELECT "film_id" FROM "film" where "film_id" <= 5
EXCEPT
SELECT "film_id" FROM "film" where "film_id" >= 4;
```

### Data Manipulation Language (DML) Statements

- INSERT – insert one or more rows into a table.
- UPDATE – update existing data in a table.
- DELETE – delete data from a table permanently.

# Reference

- SQL Tutorial https://www.sqltutorial.org/